---
title: "Lab 4 Team"
author: "Nicholas Heger"
date: "10/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(ggplot2)
defense <- read_excel('./cyclonesFootball2019.xlsx', sheet='Defensive')
offense <- read_excel('./cyclonesFootball2019.xlsx', sheet='Offensive')
biography <- read_excel('./cyclonesFootball2019.xlsx', sheet='Biography')
defense2018 <- read_excel('./cyclonesFootball2018.xlsx', sheet='Defensive')
offense2018 <- read_excel('./cyclonesFootball2018.xlsx', sheet='Offensive')
biography2018 <- read_excel('./cyclonesFootball2018.xlsx', sheet='Biography')
```

## Part 1  
1. The player names and opponent names are factors.  
```{r}
defClean <- defense
defClean$Name <- as.factor(defClean$Name)
defClean$Opponent_Opponent <- as.factor(defClean$Opponent_Opponent)
offClean <- offense
offClean$Name <- as.factor(offClean$Name)
offClean$Opponent_Opponent <- as.factor(offClean$Opponent_Opponent)
bioClean <- biography
bioClean$Name <- as.factor(bioClean$Name)
str(defClean)
str(offClean)
str(bioClean)
```
  
2. All offensive and defensive statistics are numerical. The Weight column in biography is also numeric. (Optional: Learn to mutate multiple columns systematically using the ?across function in `dplyr)  
```{r}
defClean[, 3:11] <- apply(defClean[, 3:11], 2, function(x) as.numeric(x))
offClean[, 3:12] <- apply(offClean[, 3:12], 2, function(x) as.numeric(x))
str(defClean)
str(offClean)
```
  
3. Change the Height column in biography into numeric. (Hint: Use a unit that works well)  
```{r}
bioClean <- biography %>% 
  mutate(Weight = as.numeric(Weight)) %>% 
  separate(Height, c('feet', 'inches'), sep='-') %>% 
  mutate(feet = as.numeric(feet), inches = as.numeric(inches)) %>% 
  mutate(Height_inches = (12*feet+inches)) %>% 
  select(-feet, -inches)
str(bioClean)
```
  
## Part 2  
1. Reformat the defClean data frame into a tidy format using pivot_longer. The type of statistic (Tackles_Solo, Tackles_ASST, etc) is added as a new key column named stat.  
```{r}
defClean_with_pivot <- pivot_longer(defClean, Tackles_Solo:Pass_PB, names_to = 'Defense_Skill', values_to = 'stat')
defClean_with_pivot
```
  
2. Compare the distributions of the defensive statistics. What defensive skills are rare?  
```{r}
ggplot(defClean_with_pivot, aes(x=Defense_Skill, y=stat)) + geom_bar(stat='identity')
```
  
All of the different 3 types of turnovers are rare, as seen in the above plot.  
  
3. Did ISU have better defense against Iowa or Notre Dame? Answer this question by creating a scatterplot with x- and y-axes being the number of solo tackles (of each player). A large number of solo tackles is an indicator of good defense.  
```{r}
defClean2 <- defClean_with_pivot %>% 
  filter(Opponent_Opponent == c('Iowa', 'Notre Dame'),
         Defense_Skill == 'Tackles_Solo')
ggplot(defClean2, aes(x=Name, y=stat, color=Opponent_Opponent)) + geom_point() + theme(axis.text.x = element_text(angle = 90))
```
  
The Cyclones had better defense against Notre Dame, according the the above graph showing solo tackles.  

4. Separate the Hometown column into two columns, namely the city and the state. Print the first few rows of these two columns. (Hint: look at the sep= argument of separate)  
```{r}
bioClean <- separate(bioClean, Hometown, c('city', 'state'), sep=',')
bioClean
```
  
5. How many players are there from each state?  
```{r}
table(bioClean$state)
```
  
## Part 3  

1. Find which states do the ISU defensive and offensive players come from. Create an appropriate summary.  
```{r}
biodef <- inner_join(bioClean, defClean, by='Name')
table(biodef$state)
biooff <- inner_join(bioClean, offClean, by='Name')
table(biooff$state)
```
  
2. How does the performance of quarterback Brock Purdy relate to the performance of the ISU team?  
```{r}
purdy <- offClean %>% 
  filter(Name == 'Purdy, Brock')
offClean_with_pivot <- pivot_longer(offClean, Rushing_ATT:Passing_INT, names_to = 'Offense_Skill', values_to = 'stat')
purdy_with_pivot <- pivot_longer(purdy, Rushing_ATT:Passing_INT, names_to = 'Offense_Skill', values_to = 'stat')
ggplot(offClean_with_pivot, aes(x=Offense_Skill, y=stat)) + geom_bar(stat='identity') + geom_bar(data=purdy_with_pivot, stat='identity', aes(fill = 'Purdy')) + theme(axis.text.x = element_text(angle = 90))
```
  
The above plot shows Brock Purdy's stats in pink, with the rest of the team in black. We can see that Purdy's performance in passing is much higher than the rest of the team, which is to be expected from the quaterback. Purdy also has a relatively high percentage of the team's rushing yards. The team's performance in receiving yards is also much higher than that of Purdy, also expected for a quaterback. I did not compare the performance of Brock Purdy to that of the defense because the quarterback is part of the offense.  

3. Which ISU player(s) made big progress over the last year? Compare the 2019 and the 2018 data here.  
```{r}
defClean2018 <- defense2018
defClean2018$Name <- as.factor(defClean2018$Name)
defClean2018$Opponent_Opponent <- as.factor(defClean2018$Opponent_Opponent)
offClean2018 <- offense2018
offClean2018$Name <- as.factor(offClean2018$Name)
offClean2018$Opponent_Opponent <- as.factor(offClean2018$Opponent_Opponent)
defClean2018[, 3:11] <- apply(defClean2018[, 3:11], 2, function(x) as.numeric(x))
offClean2018[, 3:12] <- apply(offClean2018[, 3:12], 2, function(x) as.numeric(x))
offClean2018_with_pivot <- pivot_longer(offClean2018, Rushing_ATT:Passing_INT, names_to = 'Offense_Skill', values_to = 'stat')
defClean2018_with_pivot <- pivot_longer(defClean2018, Tackles_Solo:Pass_PB, names_to = 'Defense_Skill', values_to = 'stat')
defClean2018_with_pivot$year='2018'
defClean_with_pivot$year='2019'
defClean2018_2019 = full_join(defClean2018_with_pivot, defClean_with_pivot)
offClean2018_with_pivot$year='2018'
offClean_with_pivot$year='2019'
offClean2018_2019 = full_join(offClean2018_with_pivot, offClean_with_pivot)
ggplot(defClean2018_2019, aes(x=Defense_Skill, y=stat, fill=year)) + geom_bar(stat='identity') + theme(axis.text.x = element_text(angle = 90)) + facet_wrap(~Name)
ggplot(offClean2018_2019, aes(x=Offense_Skill, y=stat, fill=year)) + geom_bar(stat='identity') + theme(axis.text.x = element_text(angle = 90)) + facet_wrap(~Name)
```
  
The above charts show the improvement in players from 2018 to 2019. Many players had their ups and downs. A few that stick out are Eyioma Uwazurike and Vance O'Rien on the defense improved significantly, and on the offense Brock Purdy improved significantly.